Competing Constructions for German Measure Noun Phrases: from Usage Data to Experiment

### 2017-04-14 Latex/article.tex
Minor fixes.

\subsection{Corpora, cognition, and alternations}

@@

The same is true if the sample size is large compared to the number of parameters, at least for finite-dimensional parameter models \citep[1119--1120]{Freedman1999}, a well-established result known as the Bernstein-von Mises theorem.

@@

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation, and \textit{rstanarm} \citep{rstanarm} for Markov-Chain Monte Carlo estimation (see Section~\ref{sec:rightstatistics}).
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was $n=5,063$ with $1,1344$ cases of \PGCa\ and $3,929$ cases of \NACa.
The results of the estimation are shown in Table~\ref{tab:bigtable} and in Figure~\ref{fig:fixeffs}.
The regressors with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \citet[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
Given the coding of the response variable, coefficients leaning to the positive side can be interpreted as favouring the \PGCa.

@@

Standard diagnostics for the Maximum Likelihood Estimation (MLE) show that the model quality is quite high.
Generalised variance inflation factors for the regressors were calculated to check for multicollinearity \citep{FoxMonette1992,ZuurEa2010}, and none of the corrected $\text{GVIF}^{1/2\text{df}}$ was higher than $1.6$.
Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.409$ and $R^2_c=0.495$ (see \citealp{Gries2015} for a basic introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.843$, which means a proportional reduction of error of $\lambda=0.297$.
The lemma intercepts have standard deviations of $\sigma_{\text{Measurelemma}}=0.448$ and $\sigma_{\text{Kindlemma}}=0.604$.

@@

In Section~\ref{sec:analyses}, it was also hypothesised that classes of measure nouns with a higher degree of grammaticalisation should favour the \PGCa.
The \textit{Measureclass} second-level predictor was successfully estimated ($\mpPB=0.001$).
Looking at the effect plot in Figure~\ref{fig:eff:measureattraction}, it is evident that abstract non-referential physical measure nouns (such as \textit{Gramm} `gram' or \textit{Liter} `litre') with a high degree of grammaticalisation favour the \NACa.
At the other end of the scale, nouns denoting natural portions like \textit{Haufen} `heap', \textit{Bündel} `bundle', \textit{Schluck} `gulp' favour the \PGCa.
These are referential nouns, confirming the hypothesis that it is prototypical of the PGC to contain two referential nouns, while the NAC only contains one (the kind noun).

@@

The first experiment tests preferences for constructions explicitly.
\cite{FordBresnan2013} use the \textit{split-100} task in which participants have to distribute $100$ points between the alternatives, assigning more points to more natural sounding alternative.
In essence, participants distribute a probability mass between two variants, which is intended to produce more subtle results compared to a two-alternative forced-choice task \citep{Rosenbach2013}, where participants have to choose one of two variants.
The split-100 paradigm has been criticised in \cite{ArppeJaervikivi2007}.
The criticism was reiterated in \cite{DivjakEa2016}, where they use a forced-choice task.
In \cite{VerhoevenTemme2017}, it was shown that results from forced-choice and split-100 experiments mostly converge, which is expected in the long run if a probability is turned into a binary decision.
I present a forced-choice experiment in Section~\ref{sec:exp:fc} and not a split-100 experiment, mainly because in a dry-run of the experiment, participants complained about the unnaturalness of the task of distributing a probability mass across two variants.
Participants had to chose between two sentences differing only in that one contained the \NACa\ and the other contained the \PGCa.
The analysis compares the probabilities assigned to stimuli by the corpus-derived model with the frequency with which participants chose the variants for the same stimuli.

@@

As stimuli, attested MNPs from the corpus study were used, but the sentences were radically simplified to avoid influences from contextual nuisance variables as much as possible.
The approach is also justified because according to the theoretical assessment in Section~\ref{sec:analyses}, the choice of variants depends mostly on a very local constructional context.
I sampled 16 MNPs from the corpus, and it was made sure that the simplifications and normalisations did not affect any of the regressors used in the corpus study.
In the simplified sentences, the case, number, etc. of the MNP remained the same as in the attested sentence, as well as the choice of lexical material within the MNP.
Eight sentences contained masculine or neuter kind nouns, the other eight contained feminine kind nouns.
Furthermore, in each of the masculine\slash neuter and feminine groups, four sentences originally containing the \NACa\ and four sentences originally containing the \PGCa\ were chosen.
More precisely, the sentences were sampled as \textit{highly prototypical examples} of \PGCa\ (high probability assigned by GLMM) and \NACa\ (low probability assigned by GLMM), respectively.%
\footnote{Remember from Section~\ref{sec:corpusstudies} that the model predicts the probability that the \PGCa\ is chosen over the \NACa.}
High and low probability were defined as the top and bottom 20\% of all probabilities assigned by the GLMM.
Lemmas and feature combinations were made unique within each group whenever possible.
The design is summarised in Table \ref{tab:experiment1:design}.

@@

A certain amount of the variance can be accounted for by idiosyncrasies of single sentences ($\sigma_{\text{Sentence}}=1.785$, $\sigma^{\text{Modelprediction}}_{\text{Sentence}}=5.996$, $16$ levels).%
\footnote{I use $\sigma^f_r$ to denote the standard deviation of the random intercepts for the fixed effect $f$ varying by random effect $r$.}
Also, among participants, there are clearly different preferences ($\sigma_{\text{Participant}}=0.781$, $\sigma^{\text{Modelprediction}}_{\text{Participant}}=0.484$, $24$ levels).
On the extreme sides, one participant chose the \PGCa\ in $13$ of $16$ cases, and two participants only chose it in $5$ of $16$ cases.
The regressor \textit{Modelprediction} achieves $\mpPB=0.007$ ($1,000$ replications) and is estimated at $5.408$ relative to an intercept of $-1.304$.
The confidence interval from a parametric bootstrap ($1,000$ replications, percentile method) for the regressor is acceptable but slightly large with a lower bound of $1.626$ and an upper bound of $8.397$.
Nakagawa \& Schielzeth's pseudo-coefficients of determination are $R^2_{m}=0.227$ and $R^2_{c}=0.561$, which means that over 22\% of the variance in the data can be explained by considering only the predictions from the corpus-based GLMM.
The effect display for the single fixed regressor \textit{Modelprediction} is given in Figure \ref{fig:afc:effects}.
The result is very clear.
The higher the probability of the \PGCa\ predicted from usage-data, the more often participants chose the \PGCa\ variant in the forced-choice task.
In summary, the forced-choice experiment clearly succeeded in corroborating the results from the corpus study in as much as the preferences extracted from usage data correspond to native speakers' choices.

@@

The second experiment tests preferences more implicitly.
It is expected that reading less prototypical variants (the one assigned a low probability by the corpus-derived model) in a given context and with given lexical material incurs a processing overhead for the reader (\citealp{Kaiser2013}).
In Section~\ref{sec:exp:spr}, a self-paced reading experiment is therefore presented.
In a very similar fashion, \cite{DivjakEa2016} apply the self-paced reading paradigm in the validation of corpus-based models.
The analysis compares the corpus-derived probabilities with potential lags in reading time for sentences with the preferred and the non-preferred constructions.

@@

The reading times were residualised per speaker based on the reading times of all words (not just the targets) by that speaker.
The adjective and the kind noun (\ie\ the constituents bearing the critical case markers) were used as the target region, such as the bracketed words in the example \textit{zwei Gläser} [\textit{spru\-deln\-des Wasser}] `two glasses of sparkling water'.
Outliers farther than 2 inter-quartile ranges from the mean logarithmised residual time were removed ($64$ data points), resulting in a total number of $n=1,152$ observations.
An LMM was specified with the logarithmised residual reading times as the response variable, and the probabilities derived from the corpus GLMM (\textit{Modelprediction}) as the main regressor.
Since the corpus GLMM predicts the probability of the \PGCa\ (vis-à-vis the \NACa), it is expected that reading times are positively correlated (longer reading times) with the probability if the stimulus actually contains the \NACa, and negatively correlated (shorter reading times) if the stimulus actually contains the \PGCa.
To account for this, an interaction between \textit{Modelprediction} and \textit{Construction} (levels \textit{PGCa} and \textit{NACa}) was added to the model.
Furthermore, the position ($1$--$98$) of the sentence in the individual experiment (\textit{Position} was included as a fixed effect to control for increasing reading speed during experiment runs.
Random intercepts were specified for \textit{Participant} and \textit{Item} (the $16$ sentence pairs are one \textit{Item} each).%
\footnote{I tried random slopes in order to keep the random effect structure maximal \citep{BarrEa2013}, but it was impossible to get the algorithm to converge due to the added complexity of the interaction.}

@@

Table~\ref{tab:exp:spr} shows the coefficient estimates with a 95\% parametric bootstrap confidence interval ($1,000$ replications, percentile method).
The standard deviation of the participant intercepts is $\sigma_{\text{Participant}}=0.079$ and of the item intercepts $\sigma_{\text{Item}}=0.037$.
Comparing the full model to a model without the main regressor \textit{Modelprediction} (and consequently also without the interaction with \textit{Construction}) in a PB test gives $\mpPB=0.036$.
Nakagawa and Schielzeth's pseudo-determination coefficients are $R^2_m=0.237$ and $R^2_c=0.346$.

@@

The overall model quality is thus good, and the effect plot for the main effect is shown in Figure~\ref{fig:spr:effects}.
The estimate for the sentences with \NACa\ is obviously imprecise, and no differences in reading times are observed.
There is a clearer effect in the sentences with \PGCa, which is also confirmed by the significant results from the bootstrapped confidence intervals (see Table~\ref{tab:exp:spr}) and from the PB test reported above.
The \PGCa\ brings about an increased reading time ($\beta_{\text{ConstructionPGCa}}=0.054$), which is plausible because it is the much rarer construction (see Section~\ref{sec:corpusstudies}).
However, if it occurs in a prototypical context and with prototypical lexical material, reading times drop ($\beta_{\text{ConstructionPGCa:Modelprediction}}=-0.125$).
This can be seen in the downward slope in the right panel of Figure~\ref{fig:spr:effects}.
This fits into the general picture in as much as the construction with the lower frequency might be developing towards a more sharply defined prototype.%
\footnote{In this context, it should be remembered from Section~\ref{sec:gettingdata} that even the \PGCd\ is much rarer that the \NACb\ ($17,252$ vs.\ $315,635$ occurrences in the auxiliary corpus samples).}
Conversely, the \NACa\ (like the NAC in general) might be the highly frequent default which does not incur reading time penalty, even if it is not the optimal choice in the given context and with the given lexical material.

>>>>

@@

The same is true if the sample size is large compared to the number of parameters, at least for finite-dimensional parameter models \citep[1119--1120]{Freedman1999}, a well-established result known as the \textit{Bernstein-von Mises theorem}.

@@

In this section, I report the results of fitting a multilevel model to the data using R \citep{R}, \textit{lme4} \citep{lme4} for Maximum Likelihood Estimation, and \textit{rstanarm} \citep{rstanarm} for `Bayesian' Markov-Chain Monte Carlo estimation (see Section~\ref{sec:rightstatistics}).
The purpose is to model the influence of the regressors specified in Table~\ref{tab:variables} on the probability that the \PGCa\ is chosen over the \NACa.
All regressors from Table~\ref{tab:variables} were included, and the measure lemma and the kind noun lemma were specified as varying-intercept random effects.
The sample size was $n=5,063$ with $1,1344$ cases of \PGCa\ and $3,929$ cases of \NACa.
The results of the estimation are shown in Table~\ref{tab:bigtable} and in Figure~\ref{fig:fixeffs}.
The regressors with the measure lemma as their unit of reference have no within-measure lemma variance, and the \textit{glmer} function automatically estimates them as \textit{group level predictors} (or \textit{second-level effects}), cf.\ \citet[265--269,302--304]{GelmanHill2006}.
The same goes for those listed with the kind lemma as their unit of reference.
Given the coding of the response variable, coefficients leaning to the positive side can be interpreted as favouring the \PGCa.

@@

Standard diagnostics for MLE show that the model quality is quite good.
Generalised variance inflation factors for the regressors were calculated to check for multicollinearity \citep{FoxMonette1992,ZuurEa2010}, and none of the corrected $\text{GVIF}^{1/2\text{df}}$ was higher than $1.6$.
Nakagawa \& Schielzeth's pseudo-coefficient of determination is $R_m^2=0.409$ and $R^2_c=0.495$ (see \citealp{Gries2015} for a basic introduction to these $R^2$ measures, or else \citealp{NakagawaSchielzeth2013}).
The rate of correct predictions is $0.843$, which means a proportional reduction of error of $\lambda=0.297$.
The lemma intercepts have standard deviations of $\sigma_{\text{Measurelemma}}=0.448$ and $\sigma_{\text{Kindlemma}}=0.604$.

@@

In Section~\ref{sec:analyses}, it was also hypothesised that classes of measure nouns with a higher degree of grammaticalisation should favour the \PGCa.
The \textit{Measureclass} second-level predictor was successfully estimated ($\mpPB=0.001$).
Looking at the effect plot in Figure~\ref{fig:eff:measureattraction}, it is evident that abstract non-referential physical measure nouns (such as \textit{Gramm} `gram' or \textit{Liter} `litre') with a high degree of grammaticalisation favour the \NACa.
At the other end of the scale, nouns denoting natural portions like \textit{Haufen} `heap', \textit{Bündel} `bundle', \textit{Schluck} `gulp' favour the \PGCa.
These are referential nouns, confirming the hypothesis that it is prototypical of the PGC to contain two referential nouns, while the NAC prototypically only contains one (the kind noun).

@@

The first experiment tests preferences for constructions explicitly.
\cite{FordBresnan2013} use the \textit{split-100} task in which participants have to distribute $100$ points between the alternatives, assigning more points to more natural sounding alternative.
In essence, participants distribute a probability mass between two variants, which is intended to produce more subtle results compared to a two-alternative forced-choice task \citep{Rosenbach2013}, where participants have to choose one of two variants.
The split-100 paradigm has been criticised in \cite{ArppeJaervikivi2007}.
The criticism was reiterated in \cite{DivjakEa2016}, where they use a forced-choice task.
In \cite{VerhoevenTemme2017}, it was shown that results from forced-choice and split-100 experiments mostly converge, which is expected in the long run if a probability is turned into a binary decision.
I present a forced-choice experiment in this Section and not a split-100 experiment, mainly because in a dry-run of the experiment, participants complained about the unnaturalness of the task of distributing a probability mass across two variants.
Participants had to chose between two sentences differing only in that one contained the \NACa\ and the other contained the \PGCa.
The analysis compares the probabilities assigned to stimuli by the corpus-derived model with the frequency with which participants chose the variants for the same stimuli.

@@

As stimuli, attested MNPs from the corpus study were used, but the sentences were radically simplified to avoid influences from contextual nuisance factors as much as possible.
The approach is also justified because according to the theoretical assessment in Section~\ref{sec:analyses}, the choice of variants depends mostly on a very local constructional context.
I sampled 16 MNPs from the corpus, and it was made sure that the simplifications and normalisations did not affect any of the regressors used in the corpus study.
In the simplified sentences, the case, number, etc. of the MNP remained the same as in the attested sentence, as well as the choice of lexical material within the MNP.
Eight sentences contained masculine or neuter kind nouns, the other eight contained feminine kind nouns.
Furthermore, in each of the masculine\slash neuter and feminine groups, four sentences originally containing the \NACa\ and four sentences originally containing the \PGCa\ were chosen.
More precisely, the sentences were sampled as \textit{highly prototypical examples} of \PGCa\ (high probability assigned by GLMM) and \NACa\ (low probability assigned by GLMM), respectively.%
\footnote{Remember from Section~\ref{sec:corpusstudies} that the model predicts the probability that the \PGCa\ is chosen over the \NACa.}
High and low probability were defined as the top and bottom 20\% of all probabilities assigned by the GLMM.
Lemmas and feature combinations were made unique within each group whenever possible.
The design is summarised in Table \ref{tab:experiment1:design}.

@@

A certain amount of the variance can be accounted for by idiosyncrasies of single sentences ($\sigma_{\text{Sentence}}=1.785$, $\sigma^{\text{Modelprediction}}_{\text{Sentence}}=5.996$, $16$ levels).%
\footnote{I use $\sigma^f_r$ to denote the standard deviation of the random intercepts for the fixed effect $f$ varying by random effect $r$.}
Also, among participants, there are clearly different preferences ($\sigma_{\text{Participant}}=0.781$, $\sigma^{\text{Modelprediction}}_{\text{Participant}}=0.484$, $24$ levels).
On the extreme sides, one participant chose the \PGCa\ in $13$ of $16$ cases, and two participants only chose it in $5$ of $16$ cases.
The regressor \textit{Modelprediction} achieves $\mpPB=0.007$ ($1,000$ replications) and is estimated at $5.408$ relative to an intercept of $-1.304$.
The confidence interval from a parametric bootstrap ($1,000$ replications, percentile method) for the regressor is acceptable but slightly large with a lower bound of $1.626$ and an upper bound of $8.397$.
The pseudo-coefficients of determination are $R^2_{m}=0.227$ and $R^2_{c}=0.561$, which means that over 22\% of the variance in the data can be explained by considering only the predictions from the corpus-based GLMM.
The effect display for the single fixed regressor \textit{Modelprediction} is given in Figure \ref{fig:afc:effects}.
The result is very clear.
The higher the probability of the \PGCa\ predicted from usage-data, the more often participants chose the \PGCa\ variant in the forced-choice task.
In summary, the forced-choice experiment clearly succeeded in corroborating the results from the corpus study in as much as the preferences extracted from usage data correspond to native speakers' choices.

@@

The second experiment tests preferences more implicitly.
It is expected that reading less prototypical variants (the one assigned a low probability by the corpus-derived model) in a given context and with given lexical material incurs a processing overhead for the reader (\citealp{Kaiser2013}).
In this Section, a self-paced reading experiment is therefore presented.
In a very similar fashion, \cite{DivjakEa2016} apply the self-paced reading paradigm in the validation of corpus-based models.
The analysis compares the corpus-derived probabilities with potential lags in reading time for sentences with the preferred and the non-preferred constructions.

@@

The reading times were residualised per speaker based on the reading times of all words (not just the targets) by that speaker.
The adjective and the kind noun (\ie\ the constituents bearing the critical case markers) were used as the target region, such as the bracketed words in the example \textit{zwei Gläser} [\textit{spru\-deln\-des Wasser}] `two glasses of sparkling water'.
Outliers farther than 2 inter-quartile ranges from the mean logarithmised residualised reading time were removed ($64$ data points), resulting in a total number of $n=1,152$ observations.
An LMM was specified with the logarithmised residual reading times as the response variable, and the probabilities derived from the corpus GLMM (\textit{Modelprediction}) as the main regressor.
Since the corpus GLMM predicts the probability of the \PGCa\ (vis-à-vis the \NACa), it is expected that reading times are positively correlated (longer reading times) with the probability if the stimulus actually contains the \NACa, and negatively correlated (shorter reading times) if the stimulus actually contains the \PGCa.
To account for this, an interaction between \textit{Modelprediction} and \textit{Construction} (levels \textit{PGCa} and \textit{NACa}) was added to the model.
Furthermore, the position ($1$--$98$) of the sentence in the individual experiment (\textit{Position} was included as a fixed effect to control for increasing reading speed during experiment runs.
Random intercepts were specified for \textit{Participant} and \textit{Item} (the $16$ sentence pairs are one \textit{Item} each).%
\footnote{I tried random slopes in order to keep the random effect structure maximal \citep{BarrEa2013}, but it was impossible to get the algorithm to converge due to the added complexity of the interaction.}

@@

Table~\ref{tab:exp:spr} shows the coefficient estimates with a 95\% parametric bootstrap confidence interval ($1,000$ replications, percentile method).
The standard deviation of the participant intercepts is $\sigma_{\text{Participant}}=0.079$ and of the item intercepts $\sigma_{\text{Item}}=0.037$.
Comparing the full model to a model without the main regressor \textit{Modelprediction} (and consequently also without the interaction with \textit{Construction}) in a PB test gives $\mpPB=0.036$.
The pseudo-coefficients of determination are $R^2_m=0.237$ and $R^2_c=0.346$.

@@

The overall model quality is high, and the effect plot for the main effect is shown in Figure~\ref{fig:spr:effects}.
The estimate for the sentences with \NACa\ is obviously imprecise, and no differences in reading times are observed.
There is a clearer effect in the sentences with \PGCa, which is also confirmed by the significant results from the bootstrapped confidence intervals (see Table~\ref{tab:exp:spr}) and from the PB test reported above.
The \PGCa\ brings about an increased reading time ($\beta_{\text{ConstructionPGCa}}=0.054$), which is plausible because it is the much rarer construction (see Section~\ref{sec:corpusstudies}).
However, if it occurs in a prototypical context and with prototypical lexical material, reading times drop ($\beta_{\text{ConstructionPGCa:Modelprediction}}=-0.125$).
This can be seen in the downward slope in the right panel of Figure~\ref{fig:spr:effects}.
This fits into the general picture in as much as the construction with the lower frequency might be developing towards a more sharply defined prototype.%
\footnote{In this context, it should be remembered from Section~\ref{sec:gettingdata} that even the \PGCd\ is much rarer that the \NACb\ ($17,252$ vs.\ $315,635$ occurrences in the auxiliary corpus samples).}
Conversely, the \NACa\ (like the NAC in general) might be the highly frequent default which does not incur reading time penalty, even if it is not the optimal choice in the given context and with the given lexical material.
